---
title: "R basics"
author: "Tobin Magle, Sarah Stevens"
teaching: 
exercises: 
questions:
- 
objectives:
- 
keypoints:
- 
source: Rmd
---

```{r setup, echo = FALSE, purl = FALSE, message = FALSE}
library(knitr)
library(tidyverse)
source("../bin/chunk-options.R")
```
Based on http://www.datacarpentry.org/R-ecology-lesson/

# Intro to R and RStudio

> ## Challenge Format
> The question goes here
>
> Use more carats for new lines 
>
> 
> > ## Solution Goes here
> > text like this
> > ```r
> > # r code goes here
> > ```
> {: .solution}
{: .challenge}

# Operators and functions

```{r}

```

> ## Challenge Format
> The question goes here
>
> Use more carats for new lines 
>
> 
> > ## Solution Goes here
> > text like this
> > ```r
> > # r code goes here
> > ```
> {: .solution}
{: .challenge}

# Data Types

> ## Challenge Format
> The question goes here
>
> Use more carats for new lines 
>
> 
> > ## Solution Goes here
> > text like this
> > ```r
> > # r code goes here
> > ```
> {: .solution}
{: .challenge}

# Loading data tables 


> ## Challenge Format
> The question goes here
>
> Use more carats for new lines 
>
> 
> > ## Solution Goes here
> > text like this
> > ```r
> > # r code goes here
> > ```
> {: .solution}
{: .challenge}

```{r}

```

<!--- HTML comment 
Comments
-->
